Follow-up Samples

### 2017-02-06 submission/manuscript_outline_20161024.Rmd

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease. Previous research suggests a link between fecal bacterial microbiome and CRC.   The overall objective of our study was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.  We also wanted to investigate the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models which could classify individuals as having a lesion or as before surgery.     
**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`).  Their change in FIT was also significantly different then those with carcinoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).  There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while the model to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).  There was a significant decrease in the model positive probability for the after surgery sample being either a lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`).   
**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion within our models but also causes detectable changes in the bacterial microbiome.  Further surveillance of these individuals will enable us to determine whether models such as the ones we present could be used to predict recurrence of colorectal cancer.

@@

This increased incidence of CRC in the younger population is concerning since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].  This increased risk can also carry with it an increased chance of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].  Therefore, there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those that may be at most susceptibility to reccurence.  Concurrently, there has also been a lot of interest in new areas that could have a role in disease pathogenesis, such as the gut bacterial microbiome.

There has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have now even shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantilzing link between the bacterial microbiome and CRC. However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma and whether it changes at all.

In this study we investigated what happened to the bacterial microbiome before and after surgery for indivdiuals with either adenoma or carcinoma.  Our anlaysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed before (initial) and after (follow up) surgery.  We also used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.

@@

***Previously Associated Cancer Bacteria***
We next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`), *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples. Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].  However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.

@@

## Discussion
There was no difference in alpha diversity metrics between the initial and follow up samples [Table S1].  Yet based on thetayc there was a significant difference between inital and follow up samples [Figure 1a] with carcinoma samples being more dissimilar to their initial samples then adenoma samples.  The change in FIT was also different between initial and follow up for those with adenoma or carcinoma [Figure 1b].  There was also a significant difference in the overall bacterial community structure for those in the carcinoma group [Figure 1d] before and after surgery but not for the adenoma group [Figure 1c].  Investigation of the OTU relative abundance before and after surgery found no single OTU to be significantly different [Figure S1].

@@

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.  We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being a before sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

There was no difference for the majority of models tested for differences in positive probablity based on whether chemotherapy or radiation was received.  There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.  These results would indicate that the findings described were specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occured.  First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.  Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.  Another limitation is that this study drew heavily from those with caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.  The time difference in collection of sample between adenoma and carcinoma that was observed could have affected our observed results. However, our random forest based model, specifically the lesion trained model, were still able to adjust positive probabilities for both adenoma and carcinoma to similar amounts despite this.

Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be overfit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

@@

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described [@kozich_development_2013].  The general processing workflow using mothur was as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

***Lesion Model Creation***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  The model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyperparameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.  Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyperparameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyperparameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.

***Before Sample Model Creation***
We also investigated whether a model could be created that could identify before and after surgery samples.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyperparameter was completed using the same procedure that was used to create the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being a before surgery sample.

@@

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.

***Analysis Overview***
Initial and follow up samples were analyzed for differences in alpha and beta diversity.  Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.  From here, all OTUs that were used in either model were then analyzed using a paired wilcoxson test.  We then investigated the relative abundance of specific previously associated CRC bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.  We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the before sample model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature before sample model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, a list of common OTUs were found for the two different models used.

@@

**Figure 1: General Differences between the Adenoma or Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  B) A significant difference was found between the adenoma and carcinoma group for change in FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`). C) NMDS of the intial and follow up samples for the Adenoma group.  D) NMDS of the initial and follow up samples for the Carcinoma group. For C) and D) the teal represents initial samples and the pink represents follow up samples.

**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples. There was a significant difference in initial and follow up sample for the OTUs classfied as *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).  B) Adenoma initial and follow up samples.  There were no significant differences between initial and follow up.

**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Before Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.  The blue line represents the reduced variable model using 100% of the data set.  A) Lesion model.  B) Defore sample model

@@

All authors declare that they do not have any relevent competing interests to report.

>>>>

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease. Previous research suggests a link between fecal bacterial microbiome and CRC.   The overall objective of our study was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.  We also wanted to investigate the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models which could classify individuals as having a lesion or as before surgery.     
**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`).  Their change in FIT was also significantly different then those with carcinoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).  There was no significant difference in any individual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while the model to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).  There was a significant decrease in the model positive probability for the after surgery sample being either a lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`).   
**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion within our models but also causes detectable changes in the bacterial microbiome.  Further surveillance of these individuals will enable us to determine whether models such as the ones we present could be used to predict recurrence of colorectal cancer.

@@

This increased incidence of CRC in the younger population is concerning since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].  This increased risk can also carry with it an increased chance of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].  Therefore, there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those that may be at most susceptibility to recurrence.  Concurrently, there has also been a lot of interest in new areas that could have a role in disease parthenogenesis, such as the gut bacterial microbiome.

There has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have now even shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the parthenogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC. However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma and whether it changes at all.

In this study we investigated what happened to the bacterial microbiome before and after surgery for individuals with either adenoma or carcinoma.  Our analysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest models and observed how these models as well as specific OTUs within them performed before (initial) and after (follow up) surgery.  We also used these models to look for similar important OTUs to identify the crucial OTUs for not only classifying initial and follow up samples but also lesion or normal.

@@

***Previously Associated Cancer Bacteria***
We next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`), *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no significant differences in any of these OTUs in the adenoma samples. Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].  However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.

@@

## Discussion
There was no difference in alpha diversity metrics between the initial and follow up samples [Table S1].  Yet based on thetayc there was a significant difference between initial and follow up samples [Figure 1a] with carcinoma samples being more dissimilar to their initial samples then adenoma samples.  The change in FIT was also different between initial and follow up for those with adenoma or carcinoma [Figure 1b].  There was also a significant difference in the overall bacterial community structure for those in the carcinoma group [Figure 1d] before and after surgery but not for the adenoma group [Figure 1c].  Investigation of the OTU relative abundance before and after surgery found no single OTU to be significantly different [Figure S1].

@@

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.  We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to commensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being a before sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

There was no difference for the majority of models tested for differences in positive probability based on whether chemotherapy or radiation was received.  There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.  These results would indicate that the findings described were specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the parthenogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.  Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the parthenogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.  Another limitation is that this study drew heavily from those with Caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be immediately extrapolate-able to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.  The time difference in collection of sample between adenoma and carcinoma that was observed could have affected our observed results. However, our random forest based model, specifically the lesion trained model, were still able to adjust positive probabilities for both adenoma and carcinoma to similar amounts despite this.

Another interesting outcome was that within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occurred because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true parthenogenesis of disease.

@@

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described [@kozich_development_2013].  The general processing workflow using mothur was as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screened for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

***Lesion Model Creation***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main testing and training of the model completed on a data set of `r length(rownames(metaI))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  The model included data on FIT and the bacterial microbiome. Non-binary data was checked for near zero variance and OTUs that had near zero variance were removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimization of the mtry hyper-parameter involved taking the samples and making 100 different 80/20 (train/test) splits of the data where normal and lesion were represented in the same proportion within both the whole data set and the 80/20 split.  Each of these splits were then run through 20 repeated 10-fold cross validations to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.  The default cutoff of 0.5 was used as the threshold to classify individuals as positive or negative for lesion.  The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data is created with the Random Forest model.

***Before Sample Model Creation***
We also investigated whether a model could be created that could identify before and after surgery samples.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure that was used to create the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being a before surgery sample.

@@

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.

***Analysis Overview***
Initial and follow up samples were analyzed for differences in alpha and beta diversity.  Next, differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.  From here, all OTUs that were used in either model were then analyzed using a paired Wilcoxson test.  We then investigated the relative abundance of specific previously associated CRC bacteria, specifically, OTUs that taxonomically classified to *Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*.  We wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build a reduced model and it was assessed for similarity to the original model.  We then used the before sample model to assess whether it could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature before sample model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, a list of common OTUs were found for the two different models used.

@@

**Figure 1: General Differences between the Adenoma or Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  B) A significant difference was found between the adenoma and carcinoma group for change in FIT (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`). C) NMDS of the initial and follow up samples for the Adenoma group.  D) NMDS of the initial and follow up samples for the Carcinoma group. For C) and D) the teal represents initial samples and the pink represents follow up samples.

**Figure 2: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples. There was a significant difference in initial and follow up sample for the OTUs classified as *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`).  B) Adenoma initial and follow up samples.  There were no significant differences between initial and follow up.

**Figure 3: Graph of the Receiver Operating Characteristic Curve for lesion and Before Sample Models.** The shaded areas represents the range of values of a 100 different 80/20 splits of the test set data using either all variables (grey) or reduced variable (red) models.  The blue line represents the reduced variable model using 100% of the data set.  A) Lesion model.  B) Before sample model

@@

All authors declare that they do not have any relevant competing interests to report.
